---
title: "El virus de la Corona"
author: "not only Stefano Cabras ... also thanks to statisticians at uc3m.es"
date: '`r format(Sys.time())` '
site: bookdown::bookdown_site
documentclass: book
output:
  bookdown::gitbook: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message = FALSE,warning = FALSE)
options(bitmapType='cairo')
rm(list=ls())
load(file="stanmod.RData")
library(reshape2)
library(plotly)
library(coda)
library(rstan)
library(knitr)
library(kableExtra)
```

# Prediction of * confirmed* cases for next days

Conditionally to:

  1. the observed data up to `r max(cvirus$fecha)` (only the *official confirmed cases*)
  2. assuming that these data reflect covid-19 epidemic evolution;
  3. the model detailed below;

Then below is the prediction of confirmed cases for last and next days updated at the time of this report (see above).

It also estimate the probability of observing the peak (defined as less increment in cases than the previous day). Prediction limits are at 95% of probability.

```{r,echo=FALSE}
prob.peak=function(predobs) round(mean(predobs[,2]<=predobs[,1])*100,1)
preds=extract(fit,pars="yp")$yp
preds=cbind(preds,extract(fit,pars="ypf")$ypf)
pp=c(NA)
for(i in 2:ncol(preds)) pp=c(pp,prob.peak(preds[,i:(i-1)])) 
dd=c(cvirus$fecha,cvirus$fecha[n]+(1:np))
dbres=data.frame(day=dd,yinf=apply(preds,2,quantile,p=0.025),
                 ymean=apply(preds,2,mean),ysup=apply(preds,2,quantile,p=0.975),
                 ppeak=pp,obs=c(cvirus$casos,rep(NA,np)))

kable(dbres[n+(-3:np),],col.names = c("Day","Inf.","Expected","Sup.","Prob of Peak","Obs."),digits = 0) %>%
  kable_styling("striped", full_width = F,position = "center") %>%
  column_spec(c(2,4), bold = T,italic = TRUE) %>%
  column_spec(4, bold = T, color = "white", background = "#D7261E") %>%
  column_spec(6, bold = T, color = "black", background = "yellow")

p=ggplot(dbres, aes(x=day, y=ymean,ymin=yinf,ymax=ysup,color=pp)) + 
  geom_point()+
 geom_ribbon(alpha = 0.5, colour = "yellow")+geom_point(data=dbres,aes(x=day,y=obs),color="red")+
  xlab("Day")+ylab("Cases")+
  ggtitle("Predicted and Observed confirmed cases")+labs(colour = "Peak's\nprobability (%)")+scale_color_gradient(low="blue", high="green")

ggplotly(p, tooltip = c("city"))

```

## Evolution of increments

```{r}
increment=NULL
for(i in 2:ncol(preds)) increment=cbind(increment,preds[,i]-preds[,i-1]) 
obsinc=cvirus$casos[-1]-cvirus$casos[-nrow(cvirus)]

dbinc=data.frame(day=dd[-1],yinf=apply(increment,2,quantile,p=0.025),
                 ymean=apply(increment,2,mean),ysup=apply(increment,2,quantile,p=0.975),
                 obs=c(obsinc,rep(NA,np)))

p=ggplot(dbinc, aes(x=day, y=ymean,ymin=yinf,ymax=ysup)) + 
  geom_point()+
 geom_ribbon(alpha = 0.5, colour = "yellow")+ geom_point(data=dbinc,aes(x=day,y=obs),color="red")+
  xlab("Day")+ylab("Cases")+
  ggtitle("Predicted and Observed increment of cases")

print(p)
```



## Data

Here is the DataBase from ISCII (https://covid19.isciii.es/) of only confirmed cases at the end of the Day.

These data can be smileading and the rest of analysis is subject to assuming that confirmed cases reflect evolution of covid-19 spread.

```{r}
rm(list=ls())
url <- "https://code.montera34.com:4443/numeroteca/covid19/-/raw/master/data/output/covid19-cases-uci-deaths-by-ccaa-spain-by-day-accumulated.csv"
cvirus <- read.table(url, sep = ",", header = T)
xxc=aggregate(cvirus$cases_registered,list(fecha=cvirus$date),sum,na.rm=TRUE)
xxc[,2]=c(0,xxc[-1,2]-xxc[-nrow(xxc),2])
cvirus=data.frame(fecha=as.Date(xxc[-1,1]),casos=xxc[-1,2])
cvirus=cvirus[order(cvirus$fecha),]
cvirus=na.omit(cvirus)
fecha=cvirus$fecha
n=nrow(cvirus)
```


## Model for marginal counts cases

Let $Y_t \in \mathcal{N}_0$ represents the number of cases at time (days) $t$ where $t=1$ is `r cvirus$fecha[1]`.

The fitted model is an ARMA(1,1) on the Poisson mean:

$$
\begin{aligned}
Y_t | \lambda_t & \sim  Poisson(\lambda_t), \mbox{ for } t>0\\
\log(\lambda_t) & = \omega+\alpha\log(1+y_{t-1})+\beta\log(\lambda_{t-1}), \mbox{ for } t>1\\
\alpha,\beta,\omega & \sim N(0,10) (i.i.d.)\\
log(\lambda_1) & \sim N(-99,0.001)
\end{aligned}
$$


Interpetation of parameters:

- $\omega$ is the mean number (in log scale) of infected (actually the certified infected);
- $\alpha$ is the short term component (i.e. the proportion of new infected with respect to the day before);
- $\beta$ is the long term component that represents the evolution with respect to the mean (this is analogue to posing a GARCH on Poisson counts);

The non Bayesian and *slighlty less complicated* version of this model can be found here:

https://papers.ssrn.com/sol3/papers.cfm?abstract_id=3551626

```{r,eval=FALSE}
rstan_options(auto_write = FALSE)
Sys.setenv(LOCAL_CPPFLAGS = '-march=native -mtune=native -axCORE-AVX2')
options(mc.cores = parallel::detectCores())

mod.cov ="
data {
  int<lower=2> n;// number of observations
  int<lower=2> np;// number of predicted days
  int<lower=0> y[n]; // Cases
  
}

parameters {
  real alpha;
  real beta; 
  real omega;
}

transformed parameters {
  vector[n] llambda;
  llambda[1]=-99;
  for(t in 2:n) llambda[t]=omega+alpha*log(1+y[t-1])+beta*llambda[t-1];
}

model {
  // Priors
  alpha ~ normal(0,10);
  beta ~ normal(0,10);
  omega ~ normal(0,10);
  // Likelihood
  y ~ poisson_log(llambda);

}

generated quantities {
  int<lower=0> yp[n];
  int<lower=0> ypf[np];
  real llambdaf[np];
  yp[1]=0;
  for(t in 2:n) yp[t] = poisson_rng(exp(llambda[t])); //y values predicted by the model
  llambdaf[1]=omega+alpha*log(1+y[n])+beta*llambda[n];
  ypf[1]=poisson_rng(exp(llambdaf[1]));
  for(t in 2:np){
      llambdaf[t]=omega+alpha*log(1+ypf[t-1])+beta*llambdaf[t-1];
      ypf[t]=poisson_rng(exp(llambdaf[t]));
  }
}
"
ii=list(omega=1.12,alpha=0.88,beta=0)
init_f <- function () list(ii,ii,ii,ii)
m1 <- stan_model(model_code = mod.cov)

```

Hamiltonian MC is used for obtaining the posterior:

```{r,eval=FALSE}
np=10
niter=10000
fit = sampling(m1,
           data=list(y=cvirus$casos,n=n,np=np), 
           iter=niter,chains = 4,
          init = init_f(),
          control = list(adapt_delta = 0.99,max_treedepth=50),
           seed = 17,refresh=0)
save(fit,cvirus,np,n,file="stanmod.RData")
```

## Goodness of Fit

The model is reliable if is able to predict what observed, when taking into account prediction uncertainty. Here is the predicted mean and 95% posterior credible interval (i.e. small with respect to the mean).

```{r}
load(file="stanmod.RData")
library(bayesplot)
preds=extract(fit,pars="yp")$yp
ppc_intervals(
  y = apply(preds,2,mean),
  yrep = preds,
  x = cvirus$casos,
  prob = 0.95
)+labs(
    x = "Observed Cases",
    y = "Predicted cases",
    title = "95% posterior predictive intervals \nvs Observed",
    subtitle = "by day"
  ) +
  panel_bg(fill = "gray95", color = NA) +
  grid_lines(color = "white")+geom_abline(intercept = 0,slope=1)

```

## Posterior Parameters

Posterior distributions of model paramters: mean and 95% credible intervals.

```{r}
int.par=c("omega","alpha","beta")
plot(fit,pars = int.par,ci.level=0.95,point_est="mean")
print(fit,pars = int.par)
```


# Iterative estimation of parameters and outof sample prediction.

This is interesting to monitor the evolution of covid-19.

If model is reliable, then also parameter evolution is reliable.

Further we check its out-of-sample prediction of cases since beginning of march. This is more reliable than the above goodness of fit.

```{r,eval=FALSE}
ndays=sum(cvirus$fecha>"2020-03-10")
evpars=data.frame(NULL)
windows=list(1:(n-ndays+1))
for(i in (n-ndays+2):n) windows=c(windows,list(1:i))
nw=length(windows)
for(i in 1:nw){
  fit = sampling(m1,
           data=list(y=cvirus$casos[windows[[i]]],
                     n=length(windows[[i]]),np=2),
           iter=niter,chains = 4,
            init = init_f(),
#           control = list(adapt_delta = 0.99,max_treedepth=50),
           seed = 11,refresh=0)
  post.par=extract(fit,pars=c(int.par,"ypf"))
  for(j in 1:length(int.par)){
    evpars=rbind(evpars,data.frame(day=cvirus$fecha[max(windows[[i]])],
                              value=post.par[[j]],
                              param=int.par[j]))
  }
  evpars=rbind(evpars,data.frame(day=cvirus$fecha[max(windows[[i]])],
                              value=post.par[[j+1]][,1],
                              param="outpred"))

  cat("i=",i,"/",nw," - ")
}
save(evpars,file="evpars.RData")
```

## Evolution of parameters (using data since the beginning)

```{r}
load(file="evpars.RData")
ggplotly(ggplot(evpars[evpars$param!="outpred",], aes(x=day, y=value, colour=param)) + geom_smooth()+
  geom_vline(xintercept=as.Date("2020-03-10"),linetype="dashed",color = "red", size=2)+
  geom_hline(yintercept=0)+
  xlab("Days")+  ylab("Posterior mean and 95% C.I."))
```

Reaching a peak means $\alpha_2<0$ and $\beta_2<0$, while disappear of the covid means all parameters less than 0.

## Out of sample prediction

```{r}
outpred=evpars[evpars$param=="outpred",]
outpred=aggregate(value~day,outpred,
                             function(xx) c(quantile(xx,0.025),mean(xx),quantile(xx,0.975)))
outpred=data.frame(do.call(cbind,outpred))
colnames(outpred)=c("day","inf","mean","sup")
outpred$predday=sort(unique(evpars$day))+1
ggplot(outpred, aes(x=predday, y=mean)) + geom_line() +
geom_point(data=cvirus[cvirus$fecha%in%outpred$day,],
           aes(x=fecha,y=casos),color="red")+
  geom_errorbar(aes(ymin=inf, ymax=sup), width=.2,position=position_dodge(0.05))+
  xlab("Days")+ ylab("Posterior outofsample mean and 95% C.I.")

```
